Reasoning Graphically about Quantum Computation

### 2008-02-29 papers/reasoning-graphically/paper.tex
background finished, intro started
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1389 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

Want to do reasoning about quantum computation by manipulating graphs

@@

For any arrow $f:A\to B$ in a compact closed category \catC 
its \emph{name} and \emph{coname}:
\[
\name{f} : I \to A^* \otimes B, \qquad \coname{f} : A \otimes  B^* \to I,
\]
and constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
$\coname{f} = e_A \circ (f \otimes \id{B^*})$.  Hence there are natural
isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
\catC(A\otimes B^*,I)$ making the \catC monoidally closed\footnote{In
  general compact closed categories  are models of multiplicative
  linear logic where $A \multimap B$ is defined as $A^\bot \otimes B$.}.
Further,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
\[
f^* = (\id{A^*} \otimes e_B) \circ (\id{A^*}\otimes f \otimes
\id{B^*}) \circ (d_A \otimes \id{B^*})
\]
By virtue of equations \eqref{eq:comcl1} and \eqref{eq:comcl2}, $f^{**} =
f$.  Thus $(\cdot)^*$ lifts to an involutive functor
$\catC^{\text{op}} \to \catC$,  making $\catC$ equivalent to its
opposite.

@@

@@

\begin{figure}[t]
  We make the convention that domain of an arrow is at the top of the
  picture, and its codomain is at the bottom.
  \centering
  \[
  \begin{array}{lllll}
      \id{A} \otimes \id{B^*} = \id{A \otimes B^*} = PIC 
      &\qquad&
      d_A = PIC  
      &\qquad& 
      f = PIC \\\\
      \name{f} = PIC 
      & \qquad &
      \coname{f} = PIC 
      &\qquad &
      f^* = PIC
  \end{array}
  \]
  \caption{Compact Closed Structure as Graphs}
  \label{fig:comcl-graphs}
\end{figure}

@@

\begin{theorem}
  Let \catC be a compact closed category  freely generated by some set
  of arrows $F$ and ground types $T$;  then \catC is equivalent to the
  category  of $T,F$-labelled graphs.
\end{theorem}

@@

While compact closed categories provide a suitable setting for
reasoning about quantum computation \cite{AbrCoe:CatSemQuant:2004},
freely generated structure will not suffice:  we need additional
equations.  These equations will be expressed as rewrites rules for
graphs.  In this section we will describe the generators and equations
used to reason about quantum computation, and show how some of its formal
properties lead to particular issues in for the rewriting machinery.

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects] the self dual object $Q$;
\item[Arrows] Two families of arrows:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : b
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down.  We use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  (The adjoints
for $H$, $\alpha_X$ and $\alpha_Z$ will be defined equationally.)  The
free compact closed category  is then given by all graphs formed by
composing and tensoring these basic graphs.

@@

  The $(\cdot)^\dag$ functor gives a vertical symmetry to the
  category, hence for every equation we have a second equation
  obtained by flipping the diagram upside down.  In additon, we have a
  ``colour symmetry'' so for each equation shown here gives rise to
  second obtained by exchanging the two colours.
  \begin{description}
  \item[Comonoid Laws] 
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{3em}{comonoid-assoc1}  &=\;&
      \inlinegraphic{3em}{comonoid-assoc2}
      &\qquad\qquad&
      \inlinegraphic{3em}{comonoid-unit1}  &=\;&
      \inlinegraphic{3em}{comonoid-unit2}  &=\;&
      \inlinegraphic{3em}{comonoid-unit3} 
      &\qquad\qquad&
      \inlinegraphic{3em}{comonoid-comm1}  &=\;&
      \inlinegraphic{3em}{comonoid-comm2}
    \end{array}
    \]
  \item[Isometry, Frobenius, and  Compact Structure]
    \[
    \begin{array}{cccccccccccccccc}
      \inlinegraphic{3.5em}{isometry1}  &=\;&
      \inlinegraphic{3.5em}{isometry2}  
      &\qquad\qquad &
      \inlinegraphic{3em}{frobenius1}  &=\;&
      \inlinegraphic{3em}{frobenius2}  
      &\qquad\qquad&
      \inlinegraphic{2.5em}{compact1}  &=\;&
      \inlinegraphic{1.3em}{compact2}  
    \end{array}
    \]
  \item[Abelian Unitary Group]
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{2.2em}{group1}  &:=\;&
      \inlinegraphic{2.2em}{group2}  &=\;&
      \inlinegraphic{2.2em}{group3} 
      &\qquad\qquad&
      \left( \inlinegraphic{2.2em}{group4}\!\right)^\dag  &=\;&
      \inlinegraphic{2.2em}{group5} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{group6}  &=\;&
      \inlinegraphic{3.5em}{group7}  &=\;&
      \inlinegraphic{3.5em}{group8} 
    \end{array}
    \]
  \item[Bilinearity] 
    \[
    \begin{array}{cccccccccccc}
      \inlinegraphic{3.5em}{alpha-commute1}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute2}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute3} 
    \end{array}
    \]
  \item[Bialgebra Laws] Let  $\inlinegraphic{1em}{bialgebra1} := 
    \inlinegraphic{1.5em}{bialgebra2}$;  then:
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{3.5em}{bialgebra3}  &=\;&
      \inlinegraphic{3em}{bialgebra4}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra5}  &=\;&
      \inlinegraphic{1.5em}{bialgebra6}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra7}  &=\;&
      \inlinegraphic{1.5em}{bialgebra8} 
  \end{array}
  \]
\item[Group Actions]
  \[
  \begin{array}{cccccccccccccccccccccc}
    \inlinegraphic{2.5em}{group-int1} &=\;&
    \inlinegraphic{2.5em}{group-int2} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int3} &=\;&
    \inlinegraphic{2em}{group-int4} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int5} &=\;&
    \inlinegraphic{1.5em}{group-int6} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int7} &=\;&
    \inlinegraphic{1.5em}{group-int8} 
  \end{array}
  \]
    \item[$H$ Properties]
    \[
    \begin{array}{cccccccccccccccccccc}
      \left(\inlinegraphic{1.5em}{H}\right)^\dag &=\;&
      \inlinegraphic{1.5em}{H} 
      &\qquad\qquad&
      \inlinegraphic{2.5em}{Heq1} &=\;&
      \inlinegraphic{2.5em}{Heq2}
    \end{array}
    \]
  \item[Colour Duality]
        \[
    \begin{array}{cccccccccccccccccccc}
      \inlinegraphic{2.5em}{Heq-delta} &=\;&
      \inlinegraphic{1.5em}{reddelta}
      &\qquad\qquad&
      \inlinegraphic{2em}{Heq-epsilon} &=\;&
      \inlinegraphic{1.5em}{redepsilon}
      &\qquad\qquad&
      \inlinegraphic{2.5em}{Heq-alpha} &=\;&
      \inlinegraphic{1.5em}{redalpha}
    \end{array}
    \]

@@

>>>>

Recent work in quantum computation has emphasised the use of graphical
languages motivated by the underlying logical structure of quantum
mechanics itself.  These techniques have a number of advantages over
the convential matrix-based approach to quantum mechanics, but in
particular the graph can directly represent the structure of quantum
states explicitly, including the

@@

Every arrow $f:A\to B$ in a compact closed category \catC
has a \emph{name} and \emph{coname}:
\[
\name{f} : I \to A^* \otimes B, \qquad \coname{f} : A \otimes  B^* \to I,
\]
are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
$\coname{f} = e_A \circ (f \otimes \id{B^*})$.  Hence there are natural
isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
\catC(A\otimes B^*,I)$ making \catC monoidally closed\footnote{In
  general compact closed categories  are models of multiplicative
  linear logic where $A \multimap B$ is defined as $A^\bot \otimes B$.}.
Further,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
\[
f^* = (\id{A^*} \otimes e_B) \circ (\id{A^*}\otimes f \otimes
\id{B^*}) \circ (d_A \otimes \id{B^*})
\]
By virtue of equations \eqref{eq:comcl1} and \eqref{eq:comcl2}, $f^{**} =
f$.  Thus $(\cdot)^*$ lifts to an involutive functor
$\catC^{\text{op}} \to \catC$,  making $\catC$ equivalent to its
opposite.

@@

\begin{figure}
  We make the convention that domain of an arrow is at the top of the
  picture, and its codomain is at the bottom.
  \centering
  \[
  \begin{array}{lllll}
      \id{A \otimes B^*} = \inlinegraphic{3em}{comcl-id}
      &\qquad&
      d_A = \inlinegraphic{2em}{comcl-eta}
      &\qquad& 
      e_B = \inlinegraphic{2em}{comcl-epsilon} 
      \\\\
      f = \inlinegraphic{3em}{comcl-f} 
      & \qquad &
      \name{f} = \inlinegraphic{3em}{comcl-name-f}
      &\qquad &
      f^* =  \inlinegraphic{3em}{comcl-dual-f}
  \end{array}
  \]
  \caption{Compact Closed Structure as Graphs}
  \label{fig:comcl-graphs}
\end{figure}

@@

\begin{remark}
  Given concrete graphs $f : A\to B$ and $g:B\to C$ there exists
  exactly one graph morphism  $\tau:  f \to (g\circ f)$ such that
  $\tau$ is an open embedding for the non-boundary nodes
  of $f$.   Indeed the intuition behind an open embedding, is that any
  such map picks out a subgraph which forms a well defined arrow in
  its own right.
\end{remark}

@@

\begin{theorem}
  Let \catC be a compact closed category, freely generated by some set
  of arrows $F$ and ground types $T$;  then \catC is equivalent to the
  category  of $T,F$-labelled graphs.
\end{theorem}

@@

While compact closed categories provide a suitable setting for
reasoning about quantum computation \cite{AbrCoe:CatSemQuant:2004},
freely generated structure will not suffice:  we need additional
equations.  These equations will be expressed as rewrites rules for
graphs.  In this section we will describe a set of generators and equations
used to reason about quantum computation, and show how some of its formal
properties lead to particular issues for the rewriting machinery.

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects] the self dual object $Q$;
\item[Arrows] Two families of arrows:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down.  We use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  (The adjoints
for $H$, $\alpha_X$ and $\alpha_Z$ will be defined equationally.)  The
free compact closed category is then given by all graphs formed by
composing and tensoring these basic graphs.  Note  that since $Q$ is
self dual, that is $Q = Q^*$, we use \emph{undirected} graphs.

@@

  The $(\cdot)^\dag$ functor gives a vertical symmetry to the
  category, hence for every equation we have a second equation
  obtained by flipping the diagram upside down.  In additon, we have a
  ``colour duality'': each equation shown here gives rise to
  second, which is obtained by exchanging the two colours.  (In fact
  this duality is derivable from the equations for involving $H$.)
  \begin{description}
  \item[Comonoid Laws] 
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{3em}{comonoid-assoc1}  &=\;&
      \inlinegraphic{3em}{comonoid-assoc2}
      &\qquad\qquad&
      \inlinegraphic{3em}{comonoid-unit1}  &=\;&
      \inlinegraphic{3em}{comonoid-unit2}  &=\;&
      \inlinegraphic{3em}{comonoid-unit3} 
      &\qquad\qquad&
      \inlinegraphic{3em}{comonoid-comm1}  &=\;&
      \inlinegraphic{3em}{comonoid-comm2}
    \end{array}
    \]
  \item[Isometry, Frobenius, and  Compact Structure]
    \[
    \begin{array}{cccccccccccccccc}
      \inlinegraphic{3.5em}{isometry1}  &=\;&
      \inlinegraphic{3.5em}{isometry2}  
      &\qquad\qquad &
      \inlinegraphic{3em}{frobenius1}  &=\;&
      \inlinegraphic{3em}{frobenius2}  
      &\qquad\qquad&
      \inlinegraphic{2.5em}{compact1}  &=\;&
      \inlinegraphic{1.3em}{compact2}  
    \end{array}
    \]
  \item[Abelian Unitary Group]
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{2.2em}{group1}  &:=\;&
      \inlinegraphic{2.2em}{group2}  &=\;&
      \inlinegraphic{2.2em}{group3} 
      &\qquad\qquad&
      \left( \inlinegraphic{2.2em}{group4}\!\right)^\dag  &=\;&
      \inlinegraphic{2.2em}{group5} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{group6}  &=\;&
      \inlinegraphic{3.5em}{group7}  &=\;&
      \inlinegraphic{3.5em}{group8} 
    \end{array}
    \]
  \item[Bilinearity] 
    \[
    \begin{array}{cccccccccccc}
      \inlinegraphic{3.5em}{alpha-commute1}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute2}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute3} 
    \end{array}
    \]
  \item[Bialgebra Laws] Let  $\inlinegraphic{1em}{bialgebra1} := 
    \inlinegraphic{1.5em}{bialgebra2}$;  then:
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{3.5em}{bialgebra3}  &=\;&
      \inlinegraphic{3em}{bialgebra4}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra5}  &=\;&
      \inlinegraphic{1.5em}{bialgebra6}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra7}  &=\;&
      \inlinegraphic{1.5em}{bialgebra8} 
  \end{array}
  \]
\item[Group Actions]
  \[
  \begin{array}{cccccccccccccccccccccc}
    \inlinegraphic{2.5em}{group-int1} &=\;&
    \inlinegraphic{2.5em}{group-int2} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int3} &=\;&
    \inlinegraphic{2em}{group-int4} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int5} &=\;&
    \inlinegraphic{1.5em}{group-int6} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int7} &=\;&
    \inlinegraphic{1.5em}{group-int8} 
  \end{array}
  \]
    \item[$H$ Properties]
    \[
    \begin{array}{cccccccccccccccccccc}
      \left(\inlinegraphic{1.5em}{H}\right)^\dag &=\;&
      \inlinegraphic{1.5em}{H} 
      &\qquad\qquad&
      \inlinegraphic{2.5em}{Heq1} &=\;&
      \inlinegraphic{2.5em}{Heq2}
    \end{array}
    \]
  \item[Colour Duality]
        \[
    \begin{array}{cccccccccccccccccccc}
      \inlinegraphic{2.5em}{Heq-delta} &=\;&
      \inlinegraphic{1.5em}{reddelta}
      &\qquad\qquad&
      \inlinegraphic{2em}{Heq-epsilon} &=\;&
      \inlinegraphic{1.5em}{redepsilon}
      &\qquad\qquad&
      \inlinegraphic{2.5em}{Heq-alpha} &=\;&
      \inlinegraphic{1.5em}{redalpha}
    \end{array}
    \]

@@

Consider the equations shown in Figure \ref{fig:graph-quant-eqns} which
involve only one colour:  these allow the remarkable \emph{spider theorem},
first noted in \cite{Coecke2006Quantum-Measure}, to be proved:

\begin{theorem}
  Let $G$ be a connected graph generated from $\delta_Z$, $\epsilon_Z$,
  $\alpha_Z$ and their adjoints;  then $G$ is totally determined by
  the number of inputs, the number of outputs, and the sum modulo
  $2\pi$ of the $\alpha$s which occur in it.
\end{theorem}

Hence any connected subgraph
involving nodes of only one colour, wherever it occurs in a graph, may
be collapsed to a single vertex labelled by a single value $\alpha$,
giving a ``spider'' as depicted below:
\[
\inlinegraphic{3em}{gen-spider}
\]
Conversely, a spider may be arbitrarily divided into subspiders,
provided the total in- and out-degree is preserved, along with and the
sum of the $\alpha $s.  Further,
one can derive from the  spider theorem $n$-fold versions of many of
the other rewrites in the system.

Spiders offer a very intuitive way to manipulate graphs,
and are far more compact and convenient in calculations than the
graphs built up naively 
from the generators.  However, formalising spiders requires moving from
finite graphs, where each vertex has bounded degree, and which are
subject to a finite number of rewrite rules, to a system where nodes
may have arbitrarily many edges, and there are infinitely many rewrite
rules.  The desire to retain these intuitive reasoning methods
motivates the extension from concrete graphs to \emph{graph patterns}.
These will comprise the main subject of this paper.
